---
title: "Fitting multi-rate Brownian motion models of character evolution"
author: Michael C. Grundler
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting multi-rate Brownian motion models of character evolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=7, fig.align="center")
```

This vignette will walk through a brief example of how we can use the `bm::bm.shift`
function to fit multi-rate models of character evolution to empirical data.

```{r}
library(bm)

# Data on turtle body size from
# Eastman, J.M., Alfara, M.E., Joyce, P., Hipp, A.L., and L.J. Harmon. 2011. 
# Evolution 65(12): 3578-3589
data(chelonia)

x = chelonia$dat
phy = phylo::read.newick(text=ape::write.tree(chelonia$phy))

# A named vector of class ‘numeric’ with log-transformed body size data for 
# 226 turtles
head(x)
```

As a first exploration of the data we'll simply plot the distribution
of contrasts.

```{r}
pic = bm::bm.pic(x, phy)
str(pic)
```

The first component of this list is a matrix with independent contrasts
stored in the first column and the contrast variances stored in the second.
The rownames of this matrix correspond to the node indices to which the
contrasts apply. For 'tree' class phylogenies used in bm package, node indices
are identical to the node indices of an ape 'phylo' object in 'cladewise'
node order.

The second component of the list are ancestral character state estimates.
These are not maximum likelihood estimates, but they are locally parsimonious
and made during the course of the contrast algorithm.

We'll use these results to color branches according to their
inferred mass, with warmer colors representing heavier masses.

```{r}
node.state = c(x, pic$ace)
rate.bin = findInterval(node.state, seq(min(node.state), max(node.state), length.out=33))
pal = c("#053061", "#2166AC", "#4393C3", "#92C5DE", "#D1E5F0", "#F7F7F7", 
        "#FDDBC7", "#F4A582", "#D6604D", "#B2182B", "#67001F")
edge.color = colorRampPalette(pal)(33)[rate.bin]
```

Note that the plot function invisibly returns the plotting coordinates,
which we'll store for later annotations.

```{r}
L = plot(phy, edge.color=edge.color, lwd=1.5)
title("Contrast-inferred turtle body size evolution")

u = numeric(phylo::Nnode(phy))
u[phylo::root(phy):phylo::Nnode(phy)] = abs(pic$pic[, 1]) / sqrt(pic$pic[, 2])

# Now we want to highlight the nodes that show large (standardized) contrasts
points(L[[1]][, 1], L[[1]][, 3], cex=5*u, lwd=0.5)
```

We can observe that large contrasts are definitely clustered in certain
regions of the phylogeny that is suggestive of rate heterogeneity, so let's 
fit a multi-rate model of body size evolution and see how it aligns with the
distribution of contrasts.

This is actually very simple:

```{r}
fit = bm::bm.shift(x, phy)
str(fit)
```

This returns a list with three components.

The first component is a vector of lineage-specific rates of character
evolution that have been average over multiple possible rate-shift
configurations. The second component is a set of weights assigned to
different rate-shift configurations. These weights were used to compute
the average rates in the first component. The final component is a function
that can be used to retrieve the lineage-specific rates of character
evolution associated with a particular rate-shift configuration.

Let's first look at the model-averaged rates. Note that each value in
this vector is the rate assigned to the edge leading to the node whose index 
is the same as the vector index.

```{r}
edge.rate = fit$avg.rates
```

Let's discretize the rates and then map them to a set of colors, where
warmer colors mean faster rates.

```{r}
rate.bin = findInterval(edge.rate, seq(0, max(edge.rate), length.out=33))
edge.color = colorRampPalette(
     c("#BEBEBE","#00008B","#FF0000","#FFA500","#FFD700"))(33)[rate.bin]

plot(phy, edge.color=edge.color, lwd=1.5)
title("Model-averaged rates")

# Now we want to highlight the nodes that show large (standardized) contrasts
points(L[[1]][, 1], L[[1]][, 3], cex=5*u, col=1, lwd=0.5)
```

And what you should see is that all the regions of phylogeny showing large 
contrasts are inferred to have elevated rates of evolution compared to regions
with small contrasts.


We can repeat the exact same process using individual rate-shift configurations
as well, not just the model-averaged rates. Let's look at the individual
configuration that has the highest score (highest AIC weight).

```{r}
best_index = which.max(fit$aic.weights)
```

Note that we have to be a little careful here due to different indexing
conventions between R and C. Rather than pass `best_index` directly, we need
to pass `best_index - 1`. This is why the names associated with the
`fit$aic.weights` vector are always one less than the vector position.

```{r}
edge.rate = fit$rate(best_index-1L)

# the particular rate-shift nodes can be recovered with
# attr(edge.rate, "shifts")

rate.bin = findInterval(edge.rate, seq(0, max(edge.rate), length.out=33))
edge.color = colorRampPalette(
     c("#BEBEBE","#00008B","#FF0000","#FFA500","#FFD700"))(33)[rate.bin]

plot(phy, edge.color=edge.color, lwd=1.5)
points(L[[1]][, 1], L[[1]][, 3], cex=5*u, col=1, lwd=0.5)
title("Best rate-shift configuration rates")
```

The overall pattern here is quite similar to the model-averaged pattern.

Finally, let's just plot the overall distribution of relative likelihood
scores.

```{r}
plot(fit$aic.weights, type='h', xlim=c(0, 100), las=1, ylab="Akaike weight",
    xlab="Model index")
```

Each model corresponds to a rate-shift configuration, and they are indexed by
the number of branches in the tree that belong to the background process. We
can see from this figure that there are two rate-shift configurations that
receive high support.

The weight assigned to `fit$aic.weights[1]` represents the relative likelihood
that the data were generated by a homogeneous rate process and it is a good
idea to always look at this value in relation to others before working with
model-averaged branch rates.

In this case, it is very unlikely that the data were generated by a constant
rate process:

```{r}
fit$aic.weights[best_index] / fit$aic.weights[1]
```

```{r echo=FALSE}
x = fit$aic.weights[best_index] / fit$aic.weights[1]
```

The best rate-shift configuration is roughly `r floor(log10(x))` orders of magnitude more likely.

